---
title: "FF Factors"
author: "Menghan Yuan"
date: "20/08/2024"
output: 
    bookdown::html_document2:
        toc: true
        toc_float: true
        toc_depth: 4
        number_sections: false
        css: style.css
bibliography: bibli.bib
link-citations: yes
biblio-style: apalike
---

<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>


```{r setup, include=F, echo=F}
library(knitr)
library(tidyverse)
# don't show code unless we explicitly set echo = TRUE
opts_chunk$set(warning = FALSE, echo = FALSE, message=FALSE, fig.align="center", fig.pos = "H")
```

## Descriptive analysis 

We have Fama-French pricing factor data from 2000-01 to 2024-06. 
We want to have a look at its descriptive statistics.

```{r}
data <- read_csv("data/FF_3Factors_US_monthly.csv")
get_stat <- function(x, q_list=c(0.025, 0.5, 0.975)){
    ## Return extended summary statistics with quantiles and sd.
    #    @q_list: vector of quantiles to calculate;
    x <- na.omit(x)
    c(summary(x), quantile(x, q_list), "sd"=sd(x) )
}
```


### Descriptive statistics
```{r descrptive-analysis}
apply(data[,-1], 2, get_stat)
```

### Histogram of SMB
```{r}
ggplot(data) +
    geom_histogram(aes(x=SMB, y=..density..), 
                   fill="#BDBCBC", 
                   color="black", 
                   binwidth = 2, 
                   boundary=0) +
    labs(x="Small minus big, SMB")
```

### Histogram of HML
```{r}
ggplot(data) +
    geom_histogram(aes(x=HML, y=..density..), 
                   fill="#BDBCBC", 
                   color="black", 
                   binwidth = 2, 
                   boundary=0) +
    labs(x="High minus low, HML")
```


## Risk Free Rate (RFR)

- Before 2013 NIBOR: [Norges Bank](https://www.norges-bank.no/en/topics/Statistics/Historical-monetary-statistics/)

- After 2013 Norwegian Overnight Weighted Average rate: [Norske Finansielle Referenser AS (NoRe)](https://nore-benchmarks.com)

- [Titlon](https://titlon.uit.no/tabledefs.php): 3 months Norwegian Government Bills, from the `bondindex` table, `CloseYield field`.


Financial service companies publish surveys about financial data. For instance, PwC Norge.

- Here is their annual report about RFF: <https://www.pwc.no/no/publikasjoner/risikopremien.html>

