---
title: "FF Factors"
author: "Menghan Yuan"
date: "18/09/2024"
output: 
    bookdown::html_document2:
        toc: true
        toc_float: true
        toc_depth: 4
        number_sections: false
        css: style.css
bibliography: bibli.bib
link-citations: yes
biblio-style: apalike
---

<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>


```{r setup, include=F, echo=F}
library(knitr)
library(kableExtra)
library(bookdown)
library(tidyverse)
# don't show code unless we explicitly set echo = TRUE
opts_chunk$set(warning = FALSE, echo = TRUE, message=FALSE, fig.align="center", fig.pos = "H")
## control long outputs by using eg `max.lines = 10`
hook_output_default <- knitr::knit_hooks$get('output')
truncate_to_lines <- function(x, n) {
   if (!is.null(n)) {
      x = unlist(stringr::str_split(x, '\n'))
      if (length(x) > n) {
         # truncate the output
         x = c(head(x, n), '...\n')
      }
      x = paste(x, collapse = '\n') # paste first n lines together
   }
   x
}
knitr::knit_hooks$set(output = function(x, options) {
   max.lines <- options$max.lines
   x <- truncate_to_lines(x, max.lines)
   hook_output_default(x, options)
})
```

## Descriptive analysis 

We have Fama-French pricing factor data from 2000-01 to 2024-06. 
We want to have a look at its descriptive statistics.

```{r}
data <- read_csv("data/FF_3Factors_US_monthly.csv")
get_stat <- function(x, q_list=c(0.025, 0.5, 0.975)){
    ## Return extended summary statistics with quantiles and sd.
    #    @q_list: vector of quantiles to calculate;
    x <- na.omit(x)
    c(summary(x), quantile(x, q_list), "sd"=sd(x) )
}
```


### Descriptive statistics
```{r descrptive-analysis}
apply(data[,-1], 2, get_stat)
```

### Histogram of SMB
```{r}
ggplot(data) +
    geom_histogram(aes(x=SMB, y=..density..), 
                   fill="#BDBCBC", 
                   color="black", 
                   binwidth = 2, 
                   boundary=0) +
    labs(x="Small minus big, SMB")
```

### Histogram of HML
```{r}
ggplot(data) +
    geom_histogram(aes(x=HML, y=..density..), 
                   fill="#BDBCBC", 
                   color="black", 
                   binwidth = 2, 
                   boundary=0) +
    labs(x="High minus low, HML")
```

___

## Risk Free Rate (RFR)

- Before 2013 NIBOR: [Norges Bank](https://www.norges-bank.no/en/topics/Statistics/Historical-monetary-statistics/)

- After 2013 Norwegian Overnight Weighted Average rate: [Norske Finansielle Referenser AS (NoRe)](https://nore-benchmarks.com)

- [Titlon](https://titlon.uit.no/tabledefs.php): 3 months Norwegian Government Bills, from the `bondindex` table, `CloseYield field`.


Financial service companies publish surveys about financial data. For instance, PwC Norge.

- Here is their annual report about RFF: <https://www.pwc.no/no/publikasjoner/risikopremien.html>

___

## AAPL case study

```{r}
library(quantmod)
aapl <- getSymbols("AAPL", 
           src = 'yahoo', 
           from = "2014-08-01", 
           to = "2024-09-17",   
           auto.assign = FALSE
           )
aapl
```


Candlestick Plot

```{r}
chartSeries(aapl, subset="2024-09", theme=chartTheme('white',up.col='green',dn.col='red'))
```

Calculate monthly return, $r_t = \frac{P_t-P_{t-1}}{P_{t-1}} = \frac{P_t}{P_{t-1}}-1$.

You may either use the function `quantmod::monthlyReturn()` or calculate manually.
```{r}
aapl <- aapl %>% 
    apply.monthly(last)
aapl$return <- monthlyReturn(aapl$AAPL.Adjusted, type='arithmetic')
aapl <- head(aapl, -1)
aapl
```

```{r}
# calculate manually
with(aapl, AAPL.Adjusted/lag(AAPL.Adjusted)-1)
```


Merge equity data with FF factors.
```{r}
reg_data <- aapl %>% 
    as_tibble() %>% 
    add_column(Date=index(aapl), .before = 1)
reg_data <- reg_data %>% 
    mutate(year=year(Date),
           mon=month(Date))
data <- data %>% mutate_at(vars(-Date), ~./100)
data <- data %>% 
    mutate(year=year(Date),
           mon=month(Date))
reg_data <- reg_data %>% 
    left_join(data[,-1], by=c("year","mon"))
reg_data
```

Calculate excess return.
```{r}
# calculate excess return
reg_data <- reg_data %>% 
    mutate(eRi = return-RF) %>% 
    rename(rmrf=`Mkt-RF`)
reg_data  %>% 
    select(-year,-mon) %>% 
    knitr::kable(floating.environment="sidewaystable", digits = 5, escape=F) %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = F, latex_options="scale_down") %>% 
    scroll_box(width = "100%", height = "500px")
```


Calculate annualized excess return.

$$
r^A = \left(1+\text{HPR}(T)\right)^{\frac{12}{T}}-1 = \left[\prod_{t=1}^T (1+r_{t}) \right]^{\frac{12}{T}}-1
$$

```{r}
n <- nrow(reg_data)-1 # number of months
r_A <- prod(reg_data$eRi+1, na.rm = TRUE)^{12/n}-1
r_A
```

Plot excess return and ERP.
```{r fig.keep=2}
plot_data <- xts(reg_data[,c("eRi","rmrf")], order.by = reg_data$Date)
plot_data
col_vec <- c("black", "red")
plot.xts(plot_data, col = col_vec, main = "Excess Returns on Asset and Market")
addLegend("topright", 
          legend.names = c("META", "Equity Risk Premium"), 
          lty = c(1, 1), 
          lwd = c(2, 2),
          col = col_vec,
          bg = "white",
          box.col = "white")
```


### CAPM

CAPM as a regression can be expressed as

$$
r_{t} - r_{f,t} = \alpha + \beta (r_{m,t}-r_{f,t}) + \varepsilon_{t}
$$ 
where $\alpha_i$ (intercept) and $\beta_i$ (slope) are the parameters to estimate.

We use the following data to estimate the parameters.

-   $r$ is the return of the asset;
-   $r_m$ is the return of the market portfolio.
-   $r_f$ is the risk free rate of interest; 
-   $r_t-r_{f,t}$ is the excess return on the asset;
-   $r_m-r_{f,t}$ is the Equity Risk Premium.

```{r}
capm_ml <- lm(eRi~rmrf, data=reg_data)
summary(capm_ml)
```

### FF 3-factor

The Fama-French three-factor model as a regression can be expressed as

$$
r_{t} - r_{f,t} = \alpha + \beta^{RMRF} (r_{m,t}-r_{f,t}) + \beta^{SMB}SMB_t + \beta^{HML}HML_t + \varepsilon_{t}
$$ 

```{r}
FF_ml <- lm(eRi~rmrf+SMB+HML, data=reg_data)
summary(FF_ml)
```

Merge the two models in one table.
```{r, results='asis'}
library(stargazer)
stargazer(capm_ml, FF_ml, type="html", 
          title="Regression Results for AAPL", 
          align = TRUE)
```




