---
title: "Autocorrelated Series"
author: "Menghan Yuan"
date: "Nov 10, 2024"
output: 
    bookdown::html_document2:
        toc: true
        toc_float: true
        toc_depth: 4
        number_sections: false
        css: "style.css"
bibliography: bibli.bib
link-citations: yes
biblio-style: apalike
editor_options: 
  chunk_output_type: console
---

```{r setup, include=F, echo=F}
library(knitr) # load packages
library(kableExtra)
library(tidyverse)
library(latex2exp)
library(stargazer)
library(xts)
# don't show code unless we explicitly set echo = TRUE
opts_chunk$set(echo = TRUE, message=FALSE, fig.align="center", fig.pos = "H")
opts <- options(knitr.kable.NA = "")

## control long outputs by using eg `max.lines = 10`
hook_output_default <- knitr::knit_hooks$get('output')
truncate_to_lines <- function(x, n) {
   if (!is.null(n)) {
      x = unlist(stringr::str_split(x, '\n'))
      if (length(x) > n) {
         # truncate the output
         x = c(head(x, n), '...\n')
      }
      x = paste(x, collapse = '\n') # paste first n lines together
   }
   x
}
knitr::knit_hooks$set(output = function(x, options) {
   max.lines <- options$max.lines
   x <- truncate_to_lines(x, max.lines)
   hook_output_default(x, options)
})
# add labels to multi-panel figures
put.fig.letter <- function(label, location="topleft", x=NULL, y=NULL, 
                           offset=c(0, 0), ...) {
    # offset[1]: negative to move to the left; positive to move to the right
    # offset[2]: negative to move downward; positive to move upward
    if(length(label) > 1) {
        warning("length(label) > 1, using label[1]")
    }
    if(is.null(x) | is.null(y)) {
        coords <- switch(location,
                         topleft = c(0.015,0.98),
                         topcenter = c(0.5525,0.98),
                         topright = c(0.985, 0.98),
                         bottomleft = c(0.015, 0.02), 
                         bottomcenter = c(0.5525, 0.02), 
                         bottomright = c(0.985, 0.02),
                         c(0.015, 0.98) )
    } else {
        coords <- c(x,y)
    }
    this.x <- grconvertX(coords[1] + offset[1], from="nfc", to="user")
    this.y <- grconvertY(coords[2] + offset[2], from="nfc", to="user")
    text(labels=label[1], x=this.x, y=this.y, xpd=T, ...)
}
```


# AR(1) Visualization

The first-order autoregressive process, denoted AR(1), is
$$
\varepsilon_t = \rho \varepsilon_{t-1} + w_t
$$
where $w_t$ is a strictly stationary and ergodic white noise process with 0 mean and variance $\sigma^2_w$. 

```{r echo=FALSE, include=F}
n <- 100
eff_seed <- sample(1:2^15, 1)
eff_seed <- 14375
set.seed(eff_seed)
print(sprintf("Seed for session: %s", eff_seed))
w <- rnorm(n, sd=20)
f_name <- "images/white_noise.png"
ppi <- 300
png(f_name, width=9.4*ppi, height=4.61*ppi, res=ppi)
plot(w, type='l', main=TeX("White noise, $\\sigma=20$"), 
     ylab=TeX("$w_t$"), xlab="Time", cex.lab=1.5)
dev.off()
```

```{r WN, echo=FALSE, out.width = "80%", fig.cap="White noise process with $\\sigma=20$."}
include_graphics(f_name)
```


To illustrate the behavior of the AR(1) process, Fig. \@ref(fig:AR1) plots two simulated AR(1) processes. 
Each is generated using the white noise process et displayed in Fig. \@ref(fig:WN).

The plot in Fig. \@ref(fig:AR1)(a) sets $\rho=0.5$ and the plot in Fig. \@ref(fig:AR1)(b) sets $\rho=0.95$.

Remarks

- Fig. \@ref(fig:AR1)(b) is more smooth than Fig. \@ref(fig:AR1)(a). 
- The smoothing increases with $\rho$.

```{r echo=FALSE, include=F}
# simulate AR(1)
# rho=0.5
epsilon_1 <- NULL
epsilon_1[1] <- 1
for (i in 2:n) {
  epsilon_1[i] <- 0.5*epsilon_1[i-1] + w[i]
}
# rho=0.95
epsilon_2 <- NULL
epsilon_2[1] <- 1
for (i in 2:n) {
  epsilon_2[i] <- 0.95*epsilon_2[i-1] + w[i]
}
```

```{r echo=FALSE, include=F}
f_name <- "images/autocorrelated_series.png"
my.locations <- rep("topleft", 2)
png(f_name, width=9.35*ppi, height=7.12*ppi, res=ppi)
par(mfrow=c(2,1), oma=c(0,0.5,0,2), mgp=c(2,0.8,0), xpd=T )
par(mar=c(2.5, 3.5, 2, 1))
plot(epsilon_1, type="l", main=TeX("$\\rho=0.5$"), 
     xlab=" ", ylab=TeX("$\\epsilon_t$"), cex.lab=1.5 )
put.fig.letter(label="(a)", location=my.locations[1], font=2, offset = c(0.1, -0.15))
par(mar=c(3.5, 3.5, 1, 1), xpd=T)
plot(epsilon_2, type="l", main=TeX("$\\rho=0.95$"), 
     xlab="Time", ylab=TeX("$\\epsilon_t$"), cex.lab=1.5 )
put.fig.letter(label="(b)", location=my.locations[2], font=2, offset = c(0.1, -0.1))
dev.off()
```



```{r AR1, echo=FALSE, out.width = "80%", fig.cap="Simulated AR(1) processes with positive $\\rho$. (a) $\\rho=0.5$, (b) $\\rho=0.95$. White noise uses the series in Fig.&nbsp;\\@ref(fig:WN)."}
include_graphics(f_name)
```

___

We have seen the cases when $\rho$ is positive, now let's consider when $\rho$ is negative.
Fig. \@ref(fig:AR1-neg)(a) shows an AR(1) process with $\rho=-0.5$, and Fig. \@ref(fig:AR1-neg)(a) shows an AR(1) process with $\rho=-0.95\,.$

We see that the sample path is very choppy when $\rho$ is negative. The different patterns for positive and negative $\rho$'s are due to their autocorrelation functions (ACFs).

```{r echo=FALSE}
# rho=-0.5
epsilon_3 <- NULL
epsilon_3[1] <- 1
for (i in 2:n) {
  epsilon_3[i] <- -0.5*epsilon_3[i-1] + w[i]
}

# rho=-0.95
epsilon_4 <- NULL
epsilon_4[1] <- 1
for (i in 2:n) {
  epsilon_4[i] <- -1*epsilon_4[i-1] + w[i]
}
```



```{r echo=FALSE, include=F}
f_name <- "images/autocorrelated_series_negative-rho.png"
my.locations <- rep("topleft", 2)
png(f_name, width=9.35*ppi, height=7.12*ppi, res=ppi)
par(mfrow=c(2,1), oma=c(0,0.5,0,2), mgp=c(2,0.8,0), xpd=T )
par(mar=c(2.5, 3.5, 2, 1))
plot(epsilon_3, type="l", main=TeX("$\\rho = -0.5$"), 
     xlab=" ", ylab=TeX("$\\epsilon_t$"), cex.lab=1.5 )
put.fig.letter(label="(a)", location=my.locations[1], font=2, offset = c(0.1, -0.15))
par(mar=c(3.5, 3.5, 1, 1), xpd=T)
plot(epsilon_4, type="l", main=TeX("$\\rho = -0.95$"), 
     xlab=" ", ylab=TeX("$\\epsilon_t$"), cex.lab=1.5 )
put.fig.letter(label="(b)", location=my.locations[2], font=2, offset = c(0.1, -0.1))
dev.off()
```


```{r AR1-neg, echo=FALSE, out.width = "80%", fig.cap="Simulated AR(1) processes with negtive $\\rho$. (a) $\\rho=-0.5$, (b) $\\rho=-0.95$. White noise uses the series in Fig.&nbsp;\\@ref(fig:WN)."}
include_graphics(f_name)
```


# Mathematical Properties

Let's formulate a AR(1) model as follows:

$$
\begin{align} (\#eq:AR1)
\varepsilon_t = \rho \varepsilon_{t-1} + w_t 
\end{align}
$$
where $w_t$ is a white noise series with mean zero and variance $\sigma^2_w$. We also assume $|\rho|<1$.

We can represent AR(1) model as a linear combination of the innovation $w_t$.

$$
\begin{aligned}
\varepsilon_t &= \rho \varepsilon_{t-1} + w_t 
\end{aligned}
$$















