---
title: "CAPM"
output: 
    bookdown::html_document2:
        toc: true
        toc_float: true
        toc_depth: 4
        number_sections: false
header-includes:
- \usepackage{graphicx}
- \usepackage{amsmath,amsfonts,amsthm,bm} % Math packages
---

```{r setup, include=F, echo=F}
library(knitr) # load packages
library(kableExtra)
library(tidyverse)
library(latex2exp)
library(stargazer)
# don't show code unless we explicitly set echo = TRUE
opts_chunk$set(echo = TRUE, message=FALSE, fig.align="center", fig.pos = "H")
opts <- options(knitr.kable.NA = "")

## control long outputs by using eg `max.lines = 10`
hook_output_default <- knitr::knit_hooks$get('output')
truncate_to_lines <- function(x, n) {
   if (!is.null(n)) {
      x = unlist(stringr::str_split(x, '\n'))
      if (length(x) > n) {
         # truncate the output
         x = c(head(x, n), '...\n')
      }
      x = paste(x, collapse = '\n') # paste first n lines together
   }
   x
}
knitr::knit_hooks$set(output = function(x, options) {
   max.lines <- options$max.lines
   x <- truncate_to_lines(x, max.lines)
   hook_output_default(x, options)
})
```



This script estimates CAPM for individual stocks and shows an example for 2-pass CAPM.

## CAPM for one equity

```{r block1}
f_name <- "data/NO_capm_2015-2022_monthly.csv"
data_return <- read_csv(f_name)
data_group <- data_return %>% 
    group_by(ISIN) 
data_group %>% group_keys()
groups <- data_group %>% group_split()
the_group <- groups[[1]] 
the_group %>%
    select(-rf_annualized) %>% 
    knitr::kable(floating.environment="sidewaystable", digits = 5, escape=F) %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = F, latex_options="scale_down") %>% 
    scroll_box(width = "100%", height = "500px")
```

Conduct the linear regression for one equity

```{r block2}
reg_data <- the_group %>%
    select(Return_monthly, mkt_return_monthly, rf_1month) %>%
    rename(r_i = Return_monthly,
           r_m = mkt_return_monthly,
           r_f = rf_1month) %>%
    mutate(rmrf = r_m-r_f,
           eRi = r_i-r_f)
capm_ml <- lm(eRi~rmrf, data=reg_data)
summary(capm_ml)
```

`stargazer` tabulates regression results neatly.
```{r, results='asis'}
stargazer(capm_ml, type="html", title="Regression Results")
```

## 2-pass CAPM

Get a matrix of excess returns
```{r block3}
## matrix of excess returns
library(quantmod)
f_name <- "data/NO_Rf-OSEBX_2015-2023_monthly.csv"
mkt_idx_rets <- read_csv(f_name)
mkt_idx_rets <- mkt_idx_rets %>% mutate(yrmon=as.yearmon(Date))

## convert to wide table, each row stands for time points; 
#    each col stands for equities.
data_wide <- data_return %>% 
    mutate(excess_return=Return_monthly-rf_1month) %>% 
    select(ISIN, Date, excess_return) %>% 
    spread(ISIN, excess_return) 

# merge with market index and risk-free
data_wide <- data_wide %>% 
    mutate(yrmon=as.yearmon(Date)) %>% 
    left_join(mkt_idx_rets %>% select(-Date), by="yrmon") %>% 
    select(-yrmon)

data_wide <- data_wide %>% 
    mutate(rmrf=mkt_return_monthly-rf_1month)
```


Inspect the data structure
```{r max.lines = 10}
data_wide %>% str()
```


Have a preview of *Excess Return*
```{r block5}
No_equity <- 142
ER <- data_wide[,(1:No_equity)+1] %>% as.matrix()
data_wide %>%
    knitr::kable(floating.environment="sidewaystable", digits = 5, escape=F) %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = F, latex_options="scale_down") %>% 
    scroll_box(width = "100%", height = "500px")
```


Get *Equity Risk Premium* (ERM)
```{r}
ERM <- data_wide %>% pull(rmrf)
head(ERM)
```


Get pass 2 (cross section) regression coefficients, and test if the market supports the CAPM.
```{r block6}
n <- length(ERM) 
B <- NULL
n2 <- 62
for (n2 in 61:n) {
    n1 <- n2-60
    y <- ER[n1:(n2-1),]
    x <- ERM[n1:(n2-1)]
    regr <- lm(y ~ x) # pass 1
    regr$coefficients
    betai <- regr$coefficients[2,]
    eRi <- ER[n2,]
    regr <- lm(eRi ~ betai) # pass 2
    b <- regr$coefficients
    B <- rbind(B,b)
}

head(B)
colMeans(B)
```


T-test for zero means
```{r block7}
t.test(B[,1]) 
t.test(B[,2], alternative=c("two.sided")) 
t.test(B[,2], alternative=c("greater"))
p1 <- t.test(B[,1])$p.value
p2 <- t.test(B[,2], alternative=c("greater"))$p.value 
test <- colMeans(B)
test <- rbind(test, c(p1,p2))
colnames(test)<- c("constant", "beta");
rownames(test)<- c("average", "p.value");
test
```


## Plot Security Market Line (SML)

Get $E[r_i-r_f]$ against $\beta_i$ data.
```{r}
sml_data <- cbind(eRi, betai) %>% data.frame()
sml_data <- sml_data %>% filter(betai<5)
head(sml_data)
```


Draw scatter plot
```{r}
p_sml <- ggplot(data.frame(sml_data), aes(x=betai, y=eRi)) +
    geom_point() + 
    geom_smooth(method = "lm", se=FALSE) +
    labs(x=TeX("$\\beta_i$"), y=TeX("$E[r_i-r_f]$"))
p_sml
```




