---
title: "Portfolio Optimization"
---

```{r child = "_setup.Rmd"}
```


## Agenda

- Security Market Line (SML) -- Application
- Portfolio opportunity set and plot the efficient frontier
- Optimal risky portfolio
  - Global minimum variance portfolio
  - Tangency portfolio
- Practical exercise: Portfolio optimization with R

---

class: title-slide-section, center, middle

# Security Market Line (Application)

---

## Security Market Line -- Application

CAPM:
$$\mu_i = R_f + \beta_i \; (\mu_M-R_f)$$

SML represents the linear relationship between expected return and systematic risk ($\mu_i \sim \beta_i$).

**Interpretation:**
- Intercept = risk-free rate $R_f$ ($\beta_i = 0$)
- Slope = market risk premium $(\mu_M - R_f)$
- Always passes through the market portfolio ($\beta = 1$, $\mu_M$)

**Application:**
- Point ABOVE the SML: offers a higher return than required for its beta → undervalued (underpriced)
- Point BELOW the SML: offers a lower return than required → overvalued (overpriced)

---

## Illustrative Example: Five Assets

Example Setup (monthly returns):
- $\mu_i = R_f + \beta \times (\mu_M-R_f) = 0.15\% + \beta \times 0.60\%$, Risk‑free rate $R_f = 0.15\%$, market risk premium $\mu_M-R_f = 0.60\%$.
- Observed: historical average monthly return; what you actually observed

```{r echo=FALSE, results='asis'}
rf <- 0.0015 # 0.15% monthly
mrp <- 0.0060 # 0.60% monthly expected excess market return

assets <- tibble(
  Ticker = c("KO", "DEF", "AAPL", "SLAB", "GROW"),
  Name = c("Coca-Cola Co.", "Defensive Fund", "Apple Inc.", "Silicon Labs", "Growth ETF"),
  Beta = c(0.40, 0.80, 1.30, 1.60, 1.00),
  Observed_Raw = c(0.0045, 0.0055, 0.0100, 0.0105, 0.0120) # observed average raw returns
)

assets <- assets %>%
  mutate(
    Expected_Raw = rf + Beta * mrp,
    Position = case_when(
      Observed_Raw > Expected_Raw + 1e-6 ~ "Above SML (Undervalued)",
      Observed_Raw < Expected_Raw - 1e-6 ~ "Below SML (Overvalued)",
      TRUE ~ "On SML (Fairly Valued)"
    )
  )

tbl <- assets %>%
  mutate(
    Beta = sprintf("%.2f", Beta),
    `Observed %` = sprintf("%.2f", Observed_Raw * 100),
    `Expected %` = sprintf("%.2f", Expected_Raw * 100)
  ) %>%
  select(Ticker, Name, Beta, `Observed %`, `Expected %`, Position)

tbl %>%
  kable(format = "html", align = c("l", "l", "c", "c", "c", "l"), caption = "Illustrative SML Comparison (Monthly Returns)")
```



---

## Plot: SML vs. Assets

```{r echo=FALSE, fig.width=10, fig.asp=0.5}
beta_range <- c(0, 1.7)
sml_df <- tibble(Beta = beta_range, Raw_Return = rf + Beta * mrp)

# Calculate alpha for GROW
grow_data <- assets %>% dplyr::filter(Ticker == "GROW")
grow_alpha <- grow_data$Observed_Raw - grow_data$Expected_Raw

# Calculate market return at beta = 1
market_return <- rf + 1 * mrp

p_SML <- ggplot() +
    # Add SML line
    geom_line(data = sml_df, aes(Beta, Raw_Return * 100), color = "#43418A", linewidth = 1) +
    # Observed returns
    geom_point(data = assets, aes(Beta, Observed_Raw * 100, color = Position), size = 3.2) +
    # Expected returns
    geom_point(data = assets, aes(Beta, Expected_Raw * 100), shape = 4, color = "grey40", size = 3) +
    geom_text(
        data = assets, aes(Beta, Observed_Raw * 100, label = Ticker, color = Position),
        nudge_y = 0.18,
        size = 5.5,
        show.legend = FALSE
    ) +
    # Add dashed line at beta = 1 to show market return
    geom_segment(
        aes(x = 0, xend = 1, 
        y = market_return * 100, 
        yend = market_return * 100), 
        linetype = "dashed", 
        color = "grey60", 
        alpha = 0.8) +
    # Add annotations
    annotate("text", x = -0.15, y = rf * 100 + 0.05, 
             label = sprintf("R[f] == %.2f*'%%'", rf * 100),
             hjust = 0.5, vjust = 1.4, size = 5.5, color = "#43418A", parse = TRUE) +
    annotate("text", x = -0.15, y = market_return * 100 + 0.05, 
             label = sprintf("μ[M] == %.2f*'%%'", market_return * 100),
             hjust = 0.5, vjust = 1.4, size = 5.5, color = "#43418A", parse = TRUE) +
    # Add point at market return (beta = 1)
    geom_point(aes(x = 1, y = market_return * 100), color = "#43418A", size = 2.5, shape = 16) +
    # Add point at risk free return (beta = 0)
    geom_point(aes(x = 0, y = rf * 100), color = "#43418A", size = 2.5, shape = 16) +
    # Add alpha annotation for GROW
    geom_segment(
        data = grow_data, 
        aes(x = Beta, xend = Beta, y = Expected_Raw * 100, yend = Observed_Raw * 100),
        color = "#1b7837", linetype = "dashed", linewidth = 0.8
    ) +
    annotate("text", 
           x = grow_data$Beta + 0.12, 
           y = (grow_data$Expected_Raw + grow_data$Observed_Raw) * 50, 
           label = paste("α =", sprintf("%.2f%%", grow_alpha * 100)), 
           color = "#1b7837", 
           size = 5.5, 
           hjust = 1.5,
           fontface = "italic") +
    scale_color_manual(
        values = c(
            "Above SML (Undervalued)" = "#1b7837", 
            "Below SML (Overvalued)" = "#b2182b", 
            "On SML (Fairly Valued)" = "grey40"
            )) +
    scale_x_continuous(expand = expansion(mult = c(0.1, 0.02))) + # Add padding to prevent cropping
    labs(
        title = "Security Market Line (Monthly Returns)",
        y = latex2exp::TeX("Expected Return $\\,\\mu_i$ (%)"), 
        x = latex2exp::TeX("Systematic Risk, $\\,\\beta_i$"),
        color = "Position"
    ) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "bottom")
f_name <- "images/sml_overview.png"
ggsave(f_name, p_SML, width = 10, height = 5, units = "in", dpi = 300)
```

```{r out.width="90%", echo=FALSE, fig.align='center'}
include_graphics(f_name)
```

---
## Plot: Interpretation

Notes:

- Crosses (×) mark the CAPM expected raw return for each beta. Points show observed average raw returns.
  - Green points (above line) indicate the asset offers more return than required → potential buy signal.
  - Red points (below line) indicate less return than required → potential sell / avoid.
- The distance between points and the SML represents alpha ($\alpha_i$), shown for GROW as an example.

---

## Interpreting the Example

- KO: Slightly above SML → modest undervaluation; offers return premium for low beta.
- DEF: Below SML → overvalued vs. its risk, return insufficient.
- AAPL: Above SML → delivers more than required for its high beta.
- SLAB: Slightly below SML → compensation lower than required for very high beta.
- GROW: Clearly above SML → strongest undervaluation signal.

**Discussion:**

1. Which asset you'd invest in and which asset you'd allocate away from 
2. How changes in $R_f$ or $\mu_M-R_f$ affect the SML

---
.tight-top[
## Impact of Increases in ERP
]

.tight-top.tight-bottom[
Now we increase equity risk premium ($\mu_M-R_f$), while holding $R_f$ constant.
]


```{r echo=FALSE}
# Demonstrate effect of increasing market risk premium while keeping rf constant
mrp_high <- 0.0080 # Increase to 0.8% monthly

# Create SML data for both scenarios
beta_range <- c(0, 1.7)
sml_original <- tibble(Beta = beta_range, Raw_Return = rf + Beta * mrp, Scenario = "Original (ERP = 0.6%)")
sml_high_mrp <- tibble(Beta = beta_range, Raw_Return = rf + Beta * mrp_high, Scenario = "Higher ERP (0.8%)")
sml_mrp_comparison <- bind_rows(sml_original, sml_high_mrp)
# Calculate market returns for both scenarios
market_return_original <- rf + 1 * mrp
market_return_high <- rf + 1 * mrp_high
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Create comparison plot
p_SML_mrp_comparison <- ggplot() +
    geom_line(data = sml_mrp_comparison, 
        aes(Beta, Raw_Return * 100, color = Scenario, linetype = Scenario), 
        linewidth = 1.2) +
    # Add points for original assets (only on original SML)
    geom_point(data = assets, aes(Beta, Observed_Raw * 100), color = "grey30", size = 2.5, alpha = 0.7) +
    geom_text(data = assets, 
        aes(Beta, Observed_Raw * 100, label = Ticker), 
        nudge_y = 0.15, size = 4, color = "grey30") +
    # Add market return points for both scenarios
    geom_point(aes(x = 1, y = market_return_original * 100), color = "#43418A", size = 3, shape = 16) +
    geom_point(aes(x = 1, y = market_return_high * 100), color = "#d73027", size = 3, shape = 16) +
    # Add risk-free rate point (same for both)
    geom_point(aes(x = 0, y = rf * 100), color = "black", size = 3, shape = 16) +
    annotate("text", x = 0.05, y = rf * 100 + 0.05, 
             label = sprintf("R[f] == %.2f*'%%'", rf * 100),
             hjust = 1, vjust = 0, size = 5.5, color = "black", parse = TRUE) +
    # Add dashed line at beta = 1 to show market returns
    annotate("text", x = 0.05, y = market_return_original * 100, 
             label = sprintf("μ[M-Low] == %.2f*'%%'", market_return_original * 100),
             hjust = 0.7, vjust = 0, 
             size = 5.5, color = "#43418A", 
             parse = TRUE) +
    geom_segment(
        aes(x = 0, xend = 1, 
        y = market_return_original * 100, 
        yend = market_return_original * 100), 
        linetype = "dashed", 
        color = "#43418A", 
        alpha = 0.8) +
    annotate("text", x = 0.05, y = market_return_high * 100, 
             label = sprintf("μ[M-High] == %.2f*'%%'", market_return_high * 100),
             hjust = 0.7, vjust = 0, 
             size = 5.5, color = "#d73027",
             parse = TRUE) +
    geom_segment(
        aes(x = 0, xend = 1, 
        y = market_return_high * 100, 
        yend = market_return_high * 100), 
        linetype = "dashed", 
        color = "#d73027", 
        alpha = 0.8) +
    scale_color_manual(values = c("Original (ERP = 0.6%)" = "#43418A", "Higher ERP (0.8%)" = "#d73027")) +
    scale_linetype_manual(values = c("Original (ERP = 0.6%)" = "solid", "Higher ERP (0.8%)" = "dashed")) +
    scale_x_continuous(expand = expansion(mult = c(0.15, 0.1))) +
    labs(
        title = "Effect of Increasing Equity Risk Premium on SML",
        subtitle = "Risk-free rate held constant at 0.15%",
        y = latex2exp::TeX("Expected Return $\\,\\mu_i$ (%)"), 
        x = latex2exp::TeX("Systematic Risk, $\\,\\beta_i$"),
        color = "Scenario",
        linetype = "Scenario"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position = c(0.22, 0.95),
      legend.title = element_blank()
    ) +
    guides(color = guide_legend(nrow = 1))
f_name <- "images/sml_mrp_comparison.png"
ggsave(f_name, p_SML_mrp_comparison, width = 10, height = 5, units = "in", dpi = 300)
```

```{r out.width="90%", echo=FALSE, fig.align='center'}
include_graphics(f_name)
```

---

### Impact Insights

- **Increase in equity risk premium ($\mu_M-R_f$), holding $R_f$ constant:** 
  
  Slope steepens (rotation upward around intercept at beta = 0). High-beta assets now require more return; marginal undervaluation of high-beta names can vanish; low-beta assets may appear more undervalued.

---
.tight-top[
## Impact of Increases in Risk-Free Rate
]
.tight-top.tight-bottom[
Now we increase $R_f$, while holding $\mu_M-R_f$ fixed.
]

```{r echo=FALSE, warning=FALSE, message=FALSE}
### keeping equity risk premium constant -------------------------------------
# Demonstrate effect of increasing risk-free rate while keeping equity risk premium constant
rf_high <- 0.0025 # Increase risk-free rate to 0.25%
# Keep market risk premium (equity risk premium) constant at 0.6%
mrp_constant <- 0.0060

# Create SML data for both scenarios
beta_range <- c(0, 1.7)
sml_original_rf <- tibble(Beta = beta_range, Raw_Return = rf + Beta * mrp_constant, Scenario = "Original (rf = 0.15%)")
sml_high_rf <- tibble(Beta = beta_range, Raw_Return = rf_high + Beta * mrp_constant, Scenario = "Higher rf (0.25%)")
sml_rf_comparison <- bind_rows(sml_original_rf, sml_high_rf)

# Calculate market returns for both scenarios
market_return_original_rf <- rf + 1 * mrp_constant
market_return_high_rf <- rf_high + 1 * mrp_constant

# Create comparison plot for risk-free rate change
p_SML_rf_comparison <- ggplot() +
  geom_line(
    data = sml_rf_comparison, aes(Beta, Raw_Return * 100, color = Scenario, linetype = Scenario),
    linewidth = 1.2
  ) +
  # Add points for original assets (only on original SML)
  geom_point(data = assets, aes(Beta, Observed_Raw * 100), color = "grey30", size = 2.5, alpha = 0.7) +
  geom_text(
    data = assets, aes(Beta, Observed_Raw * 100, label = Ticker),
    nudge_y = 0.15, size = 4, color = "grey30"
  ) +
  # Add market return points for both scenarios
  geom_point(aes(x = 1, y = market_return_original_rf * 100), color = "#43418A", size = 3, shape = 16) +
  geom_point(aes(x = 1, y = market_return_high_rf * 100), color = "#d73027", size = 3, shape = 16) +
  # Add risk-free rate points (different for each scenario)
  geom_point(aes(x = 0, y = rf * 100), color = "#43418A", size = 3, shape = 16) +
  geom_point(aes(x = 0, y = rf_high * 100), color = "#d73027", size = 3, shape = 16) +
  # Add parallel shift indication with arrow
  annotate("segment",
    x = 0.85, xend = 0.85,
    y = (rf + 0.85 * mrp_constant) * 100,
    yend = (rf_high + 0.85 * mrp_constant) * 100,
    arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
    color = "grey50", size = 0.8
  ) +
  annotate("text",
    x = 0.9, y = ((rf + rf_high) / 2 + 0.85 * mrp_constant) * 100,
    label = "Parallel\nshift", hjust = 0, size = 3.5, color = "grey50"
  ) +
  # Add annotations for risk-free rates
  annotate("text",
    x = 0.05, y = rf * 100 - 0.05,
    label = sprintf("r[f] == %.2f*'%%'", rf * 100),
    hjust = 0, vjust = 1, size = 4, color = "#43418A", parse = TRUE
  ) +
  annotate("text",
    x = 0.05, y = rf_high * 100 + 0.05,
    label = sprintf("r[f] == %.2f*'%%'", rf_high * 100),
    hjust = 0, vjust = 0, size = 4, color = "#d73027", parse = TRUE
  ) +
  scale_color_manual(values = c("Original (rf = 0.15%)" = "#43418A", "Higher rf (0.25%)" = "#d73027")) +
  scale_linetype_manual(values = c("Original (rf = 0.15%)" = "solid", "Higher rf (0.25%)" = "dashed")) +
  scale_x_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  labs(
    title = "Effect of Increasing Risk-Free Rate on SML",
    subtitle = "Equity risk premium held constant at 0.6%",
    y = latex2exp::TeX("Expected Return $\\,\\mu_i$ (%)"),
    x = latex2exp::TeX("Systematic Risk, $\\,\\beta_i$"),
    color = "Scenario",
    linetype = "Scenario"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = c(0.22, 0.95),
    legend.title = element_blank()
  ) +
  guides(color = guide_legend(nrow = 1))
f_name <- "images/sml_rf_comparison.png"
ggsave(f_name, p_SML_rf_comparison, width = 10, height = 5, units = "in", dpi = 300)
```

```{r out.width='90%', fig.align='center', echo=FALSE}
include_graphics(f_name)
```

---

### Impact Insights

- **Increase in $r_f$, holding $E[r_M-r_f]$ fixed:**
  
  The SML shifts up in parallel; slope stays the same. All points on the line move vertically upward by $\Delta r_f.$ No rotation occurs because the market risk premium is constant.

---

class: title-slide-section, center, middle

# Optimal Portfolio Construction

---

## Portfolio opportunity set

The collection of all feasible investment possibilities set, or *portfolio opportunity set*, in the case of two assets is simply all possible portfolios that can be formed by varying the portfolio weights such that the weights sum to one.

We summarize the expected return-risk (mean-volatility) properties of the feasible portfolios in a plot with portfolio expected return, $\mu_p$, on the vertical axis and portfolio standard deviation, $\sigma_p$, on the horizontal axis. 

Recall we have the expected return and variance of the portfolio return as follows:

$$\begin{aligned}
E[r_{p,t}] &= \sum_{i=1}^k w_{i,t} E[r_{i,t}] \\
\text{Var}[r_{p,t}] &= \sum_{i=1}^k\sum_{j=1}^k w_{i,t}\,w_{j,t}\,\text{Cov}(r_{i,t}, r_{j,t})
\end{aligned}$$

---

## Given setup for two assets A and B

```{r echo=FALSE, results='asis'}
# Create huxtable and use special characters that will be converted to math
df <- data.frame(
  Parameter = c("μ", "σ"),
  "Asset A" = c(0.175, 0.258),
  "Asset B" = c(0.055, 0.115)
)
ht <- as_hux(df)
ht <- ht %>% 
  insert_row("ρ(Asset A, Asset B)", -0.164, "", after=3) %>% 
  merge_cells(4, 2:3) %>% 
  set_bold(1, everywhere, TRUE) %>%
  set_align(everywhere, everywhere, "center")

ht %>% set_width(0.6) %>% print_html()
```

where $\mu$ is the expected return, $\sigma$ is the volatility (standard deviation) of returns, $\rho$ is the correlation coefficient between assets A and B.

---

## Equally weighted portfolio

First we consider an equally weighted portfolio $p_1$.

```{r echo=FALSE}
# initialize parameters
mu.A = 0.175
sig.A = 0.258
sig2.A = sig.A^2

mu.B = 0.055
sig.B = 0.115
sig2.B = sig.B^2

rho.AB = -0.164
sig.AB = rho.AB*sig.A*sig.B

rf <- .005 # risk-free rate
```

```{r}
# construct equally weighted ptf
x.A = 0.5
x.B = 0.5
mu.p1 = x.A*mu.A + x.B*mu.B
sig2.p1 = x.A^2 * sig2.A + x.B^2 * sig2.B + 2*x.A*x.B*sig.AB
sig.p1 = sqrt(sig2.p1)
data.tbl = t(c(mu.p1=mu.p1, sig.p1=sig.p1, sig.avg=0.5*sig.A+0.5*sig.B))
data.tbl
```

|          | Asset A | Asset B | Weight | Avg of A and B | EW ptf |
| :------: | :-----: | :-----: | ------ | -------------- | ------ |
|  $\mu$   |  0.175  |  0.055  | 0.5    | 0.115          | 0.115  |
| $\sigma$ |  0.258  |  0.115  | 0.5    | 0.187          | 0.132  |

---

## Equally weighted portfolio interpretation

This portfolio has expected return half-way between the expected returns on assets A and B, but the portfolio standard deviation is less than half-way between the asset standard deviations. This reflects risk reduction via diversification.

The equally weighted portfolio has expected return half way between the expected returns on assets A and B, but has standard deviation (volatility) that is less than half way between the standard deviations of the two assets.

For long-only portfolios, we can show that this is a general result as long as the correlation between the two assets is not perfectly positive.

---

## Long-short portfolio

Now we consider a second portfolio which allows for short selling.

> To short an asset one borrows the asset, usually from a broker, and then sells it. The proceeds from the short sale are usually kept on account with a broker and there may be restrictions on the use of these funds for the purchase of other assets. The short position is closed out when the asset is repurchased and then returned to original owner. If the asset drops in value then a gain is made on the short sale and if the asset increases in value a loss is made. 

```{r echo=FALSE, include=FALSE}
# long-short portfolio
x.A = 1.5
x.B = -0.5
mu.p2 = x.A*mu.A + x.B*mu.B
sig2.p2 = x.A^2 * sig2.A + x.B^2 * sig2.B + 2*x.A*x.B*sig.AB
sig.p2 = sqrt(sig2.p2)
data.tbl = t(c(mu.p2=mu.p2, sig.p2=sig.p2, sig.avg=1.5*sig.A-0.5*sig.B))
data.tbl
```

---

## Long-short portfolio Example

|          | Asset A | Asset B | Weight | Weighted Avg of A and B | Long-short ptf |
| :------: | :-----: | :-----: | ------ | ----------------------- | -------------- |
|  $\mu$   |  0.175  |  0.055  | 1.5    | 0.235                   | 0.235          |
| $\sigma$ |  0.258  |  0.115  | –0.5   | 0.3295                  | 0.400          |


- This portfolio has both a higher expected return and standard deviation than asset A. 
The high standard deviation is due to the short sale, which is a type of leverage, and the negative correlation between assets A and B.

- Leverage refers to an investment that is financed through borrowing. Here, the short sale of asset B produces borrowed funds used to purchase more of asset A. This increases the risk (portfolio standard deviation) of the investment.

---
.tight-top.tight-bottom[
## Portfolio opportunity set
]

.tight-top[
Now we calculate the portfolio opportunity set.
Here the portfolio weight on asset A, $x_A$, is varied from $-0.4$ to $1.4$ at a step of 0.1 and, since $x_B=1−x_A$, the weight on asset B then varies from $1.4$ to $-0.4$. 
]

```{r echo=FALSE}
x.A = seq(from=-0.4, to=1.4, by=0.1)
x.B = 1 - x.A
mu.p = x.A*mu.A + x.B*mu.B
sig2.p = x.A^2 * sig2.A + x.B^2 * sig2.B + 2*x.A*x.B*sig.AB
sig.p = sqrt(sig2.p)
port.names = paste("portfolio", 1:length(x.A), sep=" ")
data.tbl = as.data.frame(cbind(x.A, x.B, mu.p, sig.p))
rownames(data.tbl) = port.names
col.names = c('$x_{A}$',"$x_{B}$", "$\\mu_{p}$", "$\\sigma_{p}$" )
```

.pull-left.small-text[
```{r echo=FALSE}
kable(data.tbl[1:10,], col.names=col.names, excape=FALSE, digit=3)
```
]

.pull-right.small-text[
```{r echo=FALSE}
kable(data.tbl[11:19,], col.names=col.names, excape=FALSE, digit=3)
```
]

???
Portfolios 1-4 and 16-19 are the long-short portfolios and portfolios 5-15 are the long-only portfolios. 


---

## Visualizing the opportunity set

The risk-return properties of this set of feasible portfolios can be visualized using:

.pull-left[
```{r echo=FALSE, fig.width=5, fig.asp=0.8}
# mu.p ~ sig.p scatter
cex.val = 1.5
col_vec = c(rep("blue", 4), rep("black", 2),
            "green", rep("black", 8), rep("blue", 4))
par(oma=c(0,0,0,0), mar=c(3.5, 3.5, 2, 1), mgp=c(2.4,0.8,0))
plot(sig.p, mu.p, type="b", pch=16, cex=cex.val,
     ylim=c(0, max(mu.p)), xlim=c(0, max(sig.p)), 
     xlab=expression(sigma[p]), ylab=expression(mu[p]),
     cex.lab=1.5, col=col_vec)
text(x=sig.A, y=mu.A, labels="Asset A", pos=4)
text(x=sig.B, y=mu.B, labels="Asset B", pos=4)
```

- The green dot represents the *global minimum variance portfolio* (GMV). This portfolio has the property that it has the smallest variance/volatility among all feasible portfolios.
]

.pull-right[
- The black dots plus the green dot represent the long-only portfolios;
    - The black dot labeled "Asset A" is the portfolio which invests $100\%$ in asset A, i.e., ($x_A=1, x_B=0$);
    - The black dot labeled "Asset B" is the portfolio which invests $100\%$ in asset B, i.e.,  ($x_A=0, x_B=1$);
- The blue dots represent long-short portfolios.
]

---

# Global minimum variance portfolio

Computing the global minimum variance portfolio is a constrained optimization problem:

$$\begin{align*}
\underset{x_{A},x_{B}}{\min}\sigma_{p}^{2} & =x_{A}^{2}\sigma_{A}^{2}+x_{B}^{2}\sigma_{B}^{2}+2x_{A}x_{B}\sigma_{AB}\\
~s.t.~x_{A}+x_{B} & =1.
\end{align*}$$

This constrained optimization problem can be solved using two methods analytically. 

---

## Method 1: Substitution

The first method, called the *method of substitution*, uses the constraint to substitute out one of the variables to transform the constrained optimization problem in two variables into an unconstrained optimization problem in one variable. 

$$\min_{x_{A}}\sigma_{p}^{2}=x_{A}^{2}\sigma_{A}^{2}+(1-x_{A})^{2}\sigma_{B}^{2}+2x_{A}(1-x_{A})\sigma_{AB}.$$

Apply the first order condition to solve the unconstrained optimization problem.

$$0=\frac{d\sigma_{p}^{2}}{dx_{A}}=2x_{A}^{\min}\sigma_{A}^{2}-2(1-x_{A}^{\min})\sigma_{B}^{2}+2\sigma_{AB}(1-2x_{A}^{\min}).$$

We have the asset allocation for GMV portfolio:

$$\begin{equation}\begin{split}
x_{A}^{\min} &= \frac{\sigma_{B}^{2}-\sigma_{AB}}{\sigma_{A}^{2}+\sigma_{B}^{2}-2\sigma_{AB}}  \\
x_{B}^{\min} &= 1-x_{A}^{\min}.
\end{split}\end{equation}$$

---

## Method 2: Lagrange multipliers

The second method, called the method of *Lagrange multipliers*, introduces an auxiliary variable called the Lagrange multiplier and transforms the constrained optimization problem in two variables into an unconstrained optimization problem in three variables.

The Lagrangian function is formed by adding to $\sigma_p^2$ the equality constraint multiplied by an auxiliary variable $\lambda$ (the Lagrange multiplier) giving:

$$L(x_{A},x_{B},\lambda)=x_{A}^{2}\sigma_{A}^{2}+x_{B}^{2}\sigma_{B}^{2}+2x_{A}x_{B}\sigma_{AB}+\lambda(x_{A}+x_{B}-1).$$

The constrained optimum is found by solving the first-order conditions:

$$\frac{\partial L}{\partial x_A}=0,\qquad
\frac{\partial L}{\partial x_B}=0,\qquad
\frac{\partial L}{\partial\lambda}=0.$$

---

## Correlation sensitivity

The shape of the investment possibilities set is very sensitive to the correlation between assets $A$ and $B$ given the other parameters. The following figure shows the portfolio frontiers for $\rho_{AB}=-0.9, -0.5, 0, 0.5, 0.9$.

.pull-left[
<img src="https://drive.google.com/thumbnail?id=1N7wP5IfQOjSeB3omZtagmT5tzHqrOUcO&sz=w1000" alt="different rho values" style="display: block; margin-right: auto; margin-left: auto; zoom:70%; margin-top: -2em" />
]

.pull-right[
<div style="height:2em;"><br></div>
- The closer $\rho_{AB}$ is to $-1$ the more curved is the frontier toward the y-axis and the higher is the possible diversification benefit. 
- There is noticeable curvature even for positive values of  $\rho_{AB}$.
]

---

## Extreme correlation cases

<img src="https://drive.google.com/thumbnail?id=1GljEacgeeGjGl5NY_fwLptWPHPODACv4&sz=w1000" alt="perfect correlation" style="display: block; margin-right: auto; margin-left: auto; zoom:70%;" />

- If  $\rho_{AB}=1$, then the portfolio frontier approaches a straight line connecting assets A and B.
- If  $\rho_{AB}=-1$, there exists a portfolio of A and B that has positive expected return and zero variance. Meaning arbitrage opportunity.

---

## Efficient vs inefficient portfolios

.pull-left[
<img src="https://drive.google.com/thumbnail?id=1HF1Mj3sApukpi4dQ5ERbs89zqlBm3ftw&sz=w1000" alt="effiient ptf" style="display: block; margin-right: auto; margin-left: auto; zoom:70%;" />
]

.pull-right[
<div style="height:2em;"><br></div>
- Efficient portfolios are the feasible portfolios that have the highest expected return for a given level of risk as measured by portfolio standard deviation.
- Inefficient portfolios are the portfolios such that there is another feasible portfolio that has the same risk but with a lower expected return.
]

---

# Tangent portfolio

Tangent/optimal portfolio represents the portfolio which **maximizes Sharpe ratio**. 

.pull-left[
<img src="https://drive.google.com/thumbnail?id=1pTKSSwWI5Od5MrSfDqI6axOlPjVfI_zh&sz=w1000" alt="tangency ptf" style="display: block; margin-right: auto; margin-left: auto; zoom:70%;" />
]
.pull-right[
We can determine the proportions of each asset in the tangency portfolio by finding the values of  $x_A$ and $x_B$ that maximize the Sharpe ratio of a portfolio.
]

---

## Tangency optimization problem

Formally, we solve the constrained maximization problem:

$$\begin{align*}
\underset{x_{A},x_{B}}{\max}~\mathrm{SR}_{p} &= \frac{\mu_{p}-r_{f}}{\sigma_{p}}~s.t.\\
\mu_{p} & =x_{A}\mu_{A}+x_{B}\mu_{B},\\
\sigma_{p}^{2} & =x_{A}^{2}\sigma_{A}^{2}+x_{B}^{2}\sigma_{B}^{2}+2x_{A}x_{B}\sigma_{AB},\\
1 & =x_{A}+x_{B}.
\end{align*}$$

The solution to this optimization problem is given by:

$$\begin{equation}\begin{split}
x_{A}^{tan} &= \frac{(\mu_{A}-r_{f})\sigma_{B}^{2}-(\mu_{B}-r_{f})\sigma_{AB}}{(\mu_{A}-r_{f})\sigma_{B}^{2}+(\mu_{B}-r_{f})\sigma_{A}^{2}-(\mu_{A}-r_{f}+\mu_{B}-r_{f})\sigma_{AB}} \\
x_{B}^{tan} &=1-x_{A}^{tan}.
\end{split}\end{equation}$$

---

class: inverse, center, middle
# Portfolio optimization with IntroCompFinR

---

.tight-top.tight-bottom.small-text[
## Equally weighted portfolio
]

.tight-top[
- Using the same assets setup
]

.small-text[
```{r echo=FALSE, results='asis'}
ht %>% set_width(0.5) %>% print_html()
```
]

```{r echo=FALSE, message=FALSE, warning=FALSE}
# initialize parameters
suppressWarnings(library(IntroCompFinR))
mu.vec <- c(mu.A, mu.B)
names(mu.vec) <- c("Asset A", "Asset B")

sigma.vec <- c(sig.A, sig.B)
names(sigma.vec) <- c("Asset A", "Asset B")

sigma2.mat <- matrix(c(sig2.A, sig.AB, sig.AB, sig2.B), ncol=2)
colnames(sigma2.mat) <- rownames(sigma2.mat) <- c("Asset A", "Asset B")
```

- To create an equally weighted portfolio:

  .small-text[
  ```{r}
  
  k <- 2 # number of assets
  ew <- rep(1,k)/k
  equalWeight.portfolio <- getPortfolio(er=mu.vec, cov.mat=sigma2.mat, weights=ew)
  equalWeight.portfolio
  ```
  ]

  You can create any custom portfolio using `getPortfolio()` function by specifying the weights, expected return, and covariance matrix of individual assets. 


---

## Global minimum variance portfolio

The global minimum variance portfolio (allowing for short sales) is computed using:

```{r}
gmv.port <- globalMin.portfolio(mu.vec, sigma2.mat)
gmv.port
```

- If short sales are not allowed, then set `shorts = FALSE`.

---

## Efficient portfolio with target return

You can compute the portfolio with a target expected return with `efficient.portfolio`.

```{r}
target.return <- mu.vec[1]
e.port.target <- efficient.portfolio(mu.vec, sigma2.mat, target.return)
e.port.target
```

- Note that when the target return is set to the expected return of one of the assets (Asset A here), the efficient portfolio is simply that asset.

---

.tight-top.tight-bottom[
## Tangency portfolio computation
]

The tangency portfolio can be computed using `tangency.portfolio`:

.small-text[
```{r}
tan.port <- tangency.portfolio(mu.vec, sigma2.mat, risk.free = rf)
tan.port
```

```{r results='hold'}
cat("Sharpe ratio of Tangency portfolio:\n")
((tan.port$er - rf) / tan.port$sd) %>% round(4)
```
]

Compare with Sharpe ratios of individual assets:

.small-text[
```{r}
round((mu.vec - rf)/sigma.vec, 4)
```
]

---
.tight-top.tight-bottom[
## Efficient frontier
]

.tight-top.tight-bottom[
The function `efficient.frontier()` constructs the set of efficient portfolios.
]

```{r}
ef <- efficient.frontier(mu.vec, sigma2.mat, alpha.min=-0.5, alpha.max=2, nport=20)
ef
```

- ER: expected return
- SD: standard deviation
- `alpha.min` and `alpha.max` specify the range of portfolio weights on the first asset. Here, we allow the $w_A$ to vary from -0.5 to 2, corresponding to $w_b$ of 1.5 to -1.0.

---

## Plotting the efficient frontier

Use the `plot(ef)` function to create a simple plot of the efficient frontier:

```{r echo=FALSE, fig.width=8, fig.asp=0.8}
# NB: plot(ef, plot.assets = TRUE) cannot be used here. That branch of
# IntroCompFinR:::plot.Markowitz() re-evaluates the symbols stored in `ef$call`
# via eval(call$er), which resolves through the package namespace to the global
# environment -- so it only finds mu.vec/sigma2.mat in an interactive session,
# and errors with "object 'mu.vec' not found" when the deck is knitted (knitr
# evaluates chunks in render()'s `envir`, which is not globalenv). Drawing the
# same picture directly avoids the lookup entirely.
plot(ef$sd, ef$er, type = "b", col = "blue", pch = 16,
     xlim = range(c(0, sigma.vec, ef$sd)),
     ylim = range(c(0, mu.vec, ef$er)),
     xlab = "Portfolio SD", ylab = "Portfolio ER", main = "Efficient Frontier")
text(sigma.vec, mu.vec, labels = names(mu.vec))

# global min variance ptf
points(gmv.port$sd, gmv.port$er, col="green", pch=16, cex=2) 
text(gmv.port$sd, gmv.port$er, labels = "Global min", pos = 4) 

# tangency ptf
points(tan.port$sd, tan.port$er, col="red", pch=16, cex=2) 
text(tan.port$sd, tan.port$er, labels = "Tangency", pos = 3) 

# Capital Allocation Line
sr.tan = (tan.port$er - rf)/tan.port$sd 
abline(a=rf, b=sr.tan, col="green", lwd=2)
```



---

# Practical Exercise

Annual estimates of stock returns parameters for Boeing and Microsoft are given in the table below:


```{r echo=FALSE, results='asis'}
# Create huxtable and use special characters that will be converted to math
df <- data.frame(
  Parameter = c("μ", "σ²"),
  Boeing = c(0.149, 0.069),
  Microsoft = c(0.331, 0.136)
)
ht <- as_hux(df)
ht <- ht %>% 
  insert_row("ρ(Boeing, Microsoft)", -0.008, "", after=3) %>% 
  merge_cells(4, 2:3) %>% 
  set_width(0.6) %>% 
  set_bold(1, everywhere, TRUE) %>%
  set_align(everywhere, everywhere, "center")

print_html(ht)
```

Use an annual risk-free rate of $3\%$ per year for the T-bill (risk free rate). 

---

## Exercise tasks

1. Create the following portfolios.
    - Combinations of Boeing and Microsoft (with $x_{boeing}= −1, −0.9, \ldots, 2$ and $x_{msft}= 1−x_{boeing}$) 
    - Combinations of Boeing and T-bills (with $x_{boeing}=0,0.1,\ldots,2$)
    - Combinations of Microsoft and T-bills (with $x_{msft}=0,0.1,\ldots,2$). 
    - For each portfolio, compute $E(r_p)$, $\text{Var}(r_p)$ and $\text{SD}(r_p)$
    - For each portfolio, plot $E(r_p)$ vs. $\text{SD}(r_p)$ 
    - Compute the Sharpe Ratio for Boeing and Microsoft. Which asset has the highest slope value?

---

<div style="height:2em;"><br></div>
<ol start="2">
<li>Compute the global minimum variance portfolio (GMV).
  <ul>
    <li>Compute the weights of Boeing and Microsoft in the GMV portfolio using the analytical formula (1) and cross validate your results using the <code>globalMin.portfolio</code> function.</li>
    <li>Compute \(E(r_p)\), \(\text{Var}(r_p)\) and \(\text{SD}(r_p)\) for the GMV portfolio.</li>
    <li>Compute the Sharpe ratio for the GMV portfolio.</li>
  </ul>
</li>
</ol>

---

<div style="height:2em;"><br></div>
<ol start="3">
<li>Compute the tangency portfolio.
  <ul>
    <li>Compute the weights of Boeing and Microsoft in the tangency portfolio using the analytical formula (2) and cross validate your results using the <code>tangency.portfolio</code> function.</li>
    <li>Compute \(E(r_p)\), \(\text{Var}(r_p)\) and \(\text{SD}(r_p)\) for the tangency portfolio.</li>
    <li>Compute the Sharpe ratio for the tangency portfolio.</li>
  </ul>
</li>
</ol>

---

# References

Eric Zivot 2021, *Introduction to Computational Finance and Financial Econometrics with R*.

- Chapter 7: Estimation of The GWN Model: https://bookdown.org/compfinezbook/introFinRbook/Estimation-of-The-GWN-Model.html

- Chapter 12.7: Portfolio Analysis Functions in R: https://bookdown.org/compfinezbook/introcompfinr/Portfolio-Analysis-Functions.html#Portfolio-Analysis-Functions