---
title: "Portfolio Optimization Exercises"
solution: false
format: 
  pdf:
    include-in-header: ../latex/preamble.tex
    fontsize: 12pt
    number-sections: true
---

::: {=latex}
\vspace{-4em}
:::

```{r chunk-opt, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 6, fig.height = 4)
```

# Data and parameters {-}

Annual estimates of stock return parameters

|                             | Boeing | Microsoft |
| :-------------------------: | :----: | :-------: |
|           $\mu$           | 0.149  |   0.331   |
|       $\sigma^{2}$        | 0.069  |   0.136   |
| $\rho$(Boeing, Microsoft) | -0.008 |           |

Risk free rate (annual) $r_f = 0.03$.

```{r libs}
# Load libraries
library(tidyverse)
library(ggplot2)
library(data.table)
library(IntroCompFinR)
library(latex2exp)
library(knitr)
library(kableExtra)
```

```{r params-init}
# Given parameters
mu.vec <- c(0.149, 0.331)               # mu_A, mu_B
names(mu.vec) <- c("Boeing", "Microsoft")

var <- c(0.069, 0.136)              # variances
sigma <- sqrt(var)                  # standard deviations
names(sigma) <- c("Boeing", "Microsoft")

rho_AB <- -0.008
cov_AB  <- rho_AB * sigma[1] * sigma[2]

# covariance matrix
sigma2.mat <- matrix(
    c(var[1], cov_AB,
      cov_AB, var[2]),
    nrow = 2, byrow = TRUE)
colnames(sigma2.mat) <- rownames(sigma2.mat) <- c("Boeing", "Microsoft")

rf <- 0.03
```


# Create portfolios and plot $E(r_p)$ vs $\mathrm{SD}(r_p)$

We create three sets of portfolios

- Boeing and Microsoft with $x_{boeing} = -1, -0.9,\ldots, 2$ and $x_{msft} = 1 - x_{boeing}$.

- Boeing and T-bills with $x_{boeing} = 0, 0.1, \ldots, 2$ and $x_{tbill} = 1 - x_{boeing}$.

- Microsoft and T-bills with $x_{msft} = 0, 0.1, \ldots, 2$ and $x_{tbill} = 1 - x_{msft}$.

We compute portfolio expected return, variance and standard deviation for each portfolio.


```{r results='hold'}
# Helper function for portfolio stats
port_stats_two_assets <- function(
  w_a, mu_a, mu_b,
  sigma_a, sigma_b, cov_ab) {
    w_b <- 1 - w_a
    mu_p <- w_a * mu_a + w_b * mu_b
    var_p <- w_a^2 * sigma_a^2 + w_b^2 * sigma_b^2 + 2 * w_a * w_b * cov_ab
    sd_p <- sqrt(var_p)
    tibble(w_a = w_a, w_b = w_b, mu_p = mu_p, var_p = var_p, sd_p = sd_p)
}

# Boeing vs Microsoft
wA_seq <- seq(-1, 2, by = 0.1)
bm_df <- port_stats_two_assets(
    wA_seq,
    mu_a = mu.vec["Boeing"], mu_b = mu.vec["Microsoft"],
    sigma_a = sigma["Boeing"], sigma_b = sigma["Microsoft"],
    cov_ab = cov_AB
) %>%
    mutate(portfolio = "Boeing-Microsoft")

cat("Boeing-Microsoft portfolios:\n\n")
bm_df %>%
    as.data.table() %>%
    print(topn = 4)
```

```{r results='hold'}
# Boeing vs T-bills (treat tbill as mu=rf and sd=0, cov=0)
wA_seq2 <- seq(0, 2, by = 0.1)
bt_df <- tibble(
    w_a = wA_seq2,
    w_b = 1 - w_a,
    mu_p = w_a * mu.vec["Boeing"] + (1 - w_a) * rf,
    var_p = w_a^2 * var[1],
    sd_p = sqrt(var_p),
    portfolio = "Boeing-Tbill"
)

cat("Boeing-Tbill portfolios:\n\n")
bt_df %>%
    as.data.table() %>%
    print(topn = 4)
```

```{r results='hold'}
# Microsoft vs T-bills
wM_seq2 <- seq(0, 2, by = 0.1)
mt_df <- tibble(
    w_a = wM_seq2,
    w_b = 1 - w_a,
    mu_p = w_a * mu.vec["Microsoft"] + (1 - w_a) * rf,
    var_p = w_a^2 * var[2],
    sd_p = sqrt(var_p),
    portfolio = "Microsoft-Tbill"
)
cat("Microsoft-Tbill portfolios:\n\n")
mt_df %>%
    as.data.table() %>%
    print(topn = 4)
```


## Plot the mean-standard deviation frontier {-}

```{r}
# combine and prepare for plotting
all_df <- bind_rows(bm_df, bt_df, mt_df)

# Plot frontier
ggplot(all_df, aes(x = sd_p, y = mu_p, color = portfolio)) +
    geom_line() +
    geom_point(size = 1) +
    labs(
        x = TeX("SD($r_p$)"), y = TeX("E($r_p$)"),
        title = "Mean-Standard Deviation Frontier"
    ) +
    scale_color_manual(values = c("Boeing-Microsoft" = "#EE0000FF", "Boeing-Tbill" = "#3B4992FF", "Microsoft-Tbill" = "#008B45FF")) +
    theme_minimal() +
    theme(legend.position = "bottom")
```

## Sharpe Ratios of Individual Assets {-}
```{r}
# Sharpe ratios of individual assets
sharpe_assets <- (mu.vec - rf) / sigma
```

```{r echo=FALSE}
sharpe_assets %>%
    enframe(name = "Asset", value = "Sharpe") %>%
    knitr::kable(digits = 4) %>%
    column_spec(c(1, 2), width = "3cm") 
```

# Global Minimum Variance Portfolio (GMV)

```{r}
# GMV portfolio using IntroCompFinR
gmv.port <- globalMin.portfolio(er = mu.vec, cov.mat = sigma2.mat)
gmv.port
```

```{r results='hold'}
cat("Sharpe ratio of GMV portfolio:\n")
((gmv.port$er - rf) / gmv.port$sd) %>% round(4)
```

# Tangency Portfolio

```{r}
# GMV portfolio using IntroCompFinR
tan.port <- tangency.portfolio(er = mu.vec, cov.mat = sigma2.mat, risk.free = rf)
tan.port
```

```{r results='hold'}
cat("Sharpe ratio of Tangency portfolio:\n")
((tan.port$er - rf) / tan.port$sd) %>% round(4)
```