---
title: "Practical Exercises: Portfolio Optimization"
number_sections: true
output: pdf_document
---


```{r chunk-opt, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 6, fig.height = 4)
```

```{r}
# Install IntroCompFinR package using GitHub mirror
remotes::install_github("r-forge/introcompfinr", subdir = "pkg/IntroCompFinR", upgrade = "never")
```

```{r libs}
# Load libraries
library(tidyverse)
library(ggplot2)
library(data.table)
library(IntroCompFinR)
library(latex2exp)
library(knitr)
library(kableExtra)
```

# Class Example: Portfolio Optimization with `IntroCompFinR`

## Given setup for two assets A and B {-}


```{=latex} 
\begin{table}[htbp]
\centering
\begin{tabular}{lcc}
\hline
 & Asset A & Asset B \\
\hline
$\mu$    & 0.175 & 0.055 \\
$\sigma$ & 0.258 & 0.115 \\
$\rho(\text{Asset A},\text{Asset B})$ & \multicolumn{2}{c}{$-0.164$} \\
\hline
\end{tabular}
\end{table}
```

where $\mu$ is the expected return, $\sigma$ is the volatility (standard
deviation) of returns, and $\rho$ is the correlation coefficient between
assets A and B.

Risk free rate $r_f = 0.005$.

```{r class-params}
# initialize parameters
mu.A = 0.175
sig.A = 0.258
sig2.A = sig.A^2

mu.B = 0.055
sig.B = 0.115
sig2.B = sig.B^2

rho.AB = -0.164
sig.AB = rho.AB*sig.A*sig.B

rf <- .005 # risk-free rate

# IntroCompFinR takes the expected returns as a named vector and the
# covariances as a named matrix
mu.vec <- c(mu.A, mu.B)
names(mu.vec) <- c("Asset A", "Asset B")

sigma.vec <- c(sig.A, sig.B)
names(sigma.vec) <- c("Asset A", "Asset B")

sigma2.mat <- matrix(c(sig2.A, sig.AB, sig.AB, sig2.B), ncol=2)
colnames(sigma2.mat) <- rownames(sigma2.mat) <- c("Asset A", "Asset B")
```

## Equally weighted portfolio

To create an equally weighted portfolio:

```{r}
k <- 2 # number of assets
ew <- rep(1,k)/k
equalWeight.portfolio <- getPortfolio(er=mu.vec, cov.mat=sigma2.mat, weights=ew)
equalWeight.portfolio
```

You can create any custom portfolio using the `getPortfolio()` function by
specifying the weights, expected return, and covariance matrix of individual
assets.

## Global minimum variance portfolio

The global minimum variance portfolio (allowing for short sales) is computed
using:

```{r}
gmv.port <- globalMin.portfolio(mu.vec, sigma2.mat)
gmv.port
```

- If short sales are not allowed, then set `shorts = FALSE`.

## Efficient portfolio with target return

You can compute the portfolio with a target expected return with
`efficient.portfolio()`.

```{r}
target.return <- mu.vec[1]
e.port.target <- efficient.portfolio(mu.vec, sigma2.mat, target.return)
e.port.target
```

- Note that when the target return is set to the expected return of one of the
  assets (Asset A here), the efficient portfolio is simply that asset.

## Tangency portfolio

The tangency portfolio can be computed using `tangency.portfolio()`:

```{r}
tan.port <- tangency.portfolio(mu.vec, sigma2.mat, risk.free = rf)
tan.port
```

```{r results='hold'}
cat("Sharpe ratio of Tangency portfolio:\n")
((tan.port$er - rf) / tan.port$sd) %>% round(4)
```

Compare with the Sharpe ratios of the individual assets:

```{r}
round((mu.vec - rf)/sigma.vec, 4)
```

## Efficient frontier

The function `efficient.frontier()` constructs the set of efficient portfolios.

```{r}
ef <- efficient.frontier(mu.vec, sigma2.mat, alpha.min=-0.5, alpha.max=2, nport=20)
ef
```

- ER: expected return
- SD: standard deviation
- `alpha.min` and `alpha.max` specify the range of portfolio weights on
  individual assets.

## Plotting the efficient frontier

```{r fig.width=6, fig.asp=0.8}
plot(ef$sd, ef$er, type = "b", col = "blue", pch = 16,
     xlim = range(c(0, sigma.vec, ef$sd)),
     ylim = range(c(0, mu.vec, ef$er)),
     xlab = "Portfolio SD", ylab = "Portfolio ER", main = "Efficient Frontier")
text(sigma.vec, mu.vec, labels = names(mu.vec))

# global min variance ptf
points(gmv.port$sd, gmv.port$er, col="green", pch=16, cex=2)
text(gmv.port$sd, gmv.port$er, labels = "Global min", pos = 4)

# tangency ptf
points(tan.port$sd, tan.port$er, col="red", pch=16, cex=2)
text(tan.port$sd, tan.port$er, labels = "Tangency", pos = 3)

# Capital Allocation Line
sr.tan = (tan.port$er - rf)/tan.port$sd
abline(a=rf, b=sr.tan, col="green", lwd=2)
```


# Practical Exercise: Boeing and Microsoft

## Data and parameters {-}

Annual estimates of stock return parameters

```{=latex}
\begin{table}[htbp]
\centering
\begin{tabular}{lcc}
\hline
 & Boeing & Microsoft \\
\hline
$\mu$    & 0.149  & 0.331 \\
$\sigma^{2}$ & 0.069  & 0.136 \\
$\rho(\text{Boeing},\text{Microsoft})$ & \multicolumn{2}{c}{$-0.008$} \\
\hline
\end{tabular}
\end{table}
```

Risk free rate (annual) $r_f = 0.03$.

```{r params-init}
# Given parameters
mu.vec <- c(0.149, 0.331)               # mu_A, mu_B
names(mu.vec) <- c("Boeing", "Microsoft")

var <- c(0.069, 0.136)              # variances
sigma <- sqrt(var)                  # standard deviations
names(sigma) <- c("Boeing", "Microsoft")

rho_AB <- -0.008
cov_AB  <- rho_AB * sigma[1] * sigma[2]

# covariance matrix
sigma2.mat <- matrix(
    c(var[1], cov_AB,
      cov_AB, var[2]),
    nrow = 2, byrow = TRUE)
colnames(sigma2.mat) <- rownames(sigma2.mat) <- c("Boeing", "Microsoft")

rf <- 0.03
```


## Create portfolios and plot $E(r_p)$ vs $\mathrm{SD}(r_p)$

We create three sets of portfolios

- Boeing and Microsoft with $x_{boeing} = -1, -0.9,\ldots, 2$ and $x_{msft} = 1 - x_{boeing}$.

- Boeing and T-bills with $x_{boeing} = 0, 0.1, \ldots, 2$ and $x_{tbill} = 1 - x_{boeing}$.

- Microsoft and T-bills with $x_{msft} = 0, 0.1, \ldots, 2$ and $x_{tbill} = 1 - x_{msft}$.

We compute portfolio expected return, variance and standard deviation for each portfolio.


```{r results='hold'}
# Helper function for portfolio stats
port_stats_two_assets <- function(
  w_a, mu_a, mu_b,
  sigma_a, sigma_b, cov_ab) {
    w_b <- 1 - w_a
    mu_p <- w_a * mu_a + w_b * mu_b
    var_p <- w_a^2 * sigma_a^2 + w_b^2 * sigma_b^2 + 2 * w_a * w_b * cov_ab
    sd_p <- sqrt(var_p)
    tibble(w_a = w_a, w_b = w_b, mu_p = mu_p, var_p = var_p, sd_p = sd_p)
}

# Boeing vs Microsoft
wA_seq <- seq(-1, 2, by = 0.1)
bm_df <- port_stats_two_assets(
    wA_seq,
    mu_a = mu.vec["Boeing"], mu_b = mu.vec["Microsoft"],
    sigma_a = sigma["Boeing"], sigma_b = sigma["Microsoft"],
    cov_ab = cov_AB
) %>%
    mutate(portfolio = "Boeing-Microsoft")

cat("Boeing-Microsoft portfolios:\n\n")
bm_df %>%
    as.data.table() %>%
    print(topn = 4)
```

```{r results='hold'}
# Boeing vs T-bills (treat tbill as mu=rf and sd=0, cov=0)
wA_seq2 <- seq(0, 2, by = 0.1)
bt_df <- tibble(
    w_a = wA_seq2,
    w_b = 1 - w_a,
    mu_p = w_a * mu.vec["Boeing"] + (1 - w_a) * rf,
    var_p = w_a^2 * var[1],
    sd_p = sqrt(var_p),
    portfolio = "Boeing-Tbill"
)

cat("Boeing-Tbill portfolios:\n\n")
bt_df %>%
    as.data.table() %>%
    print(topn = 4)
```

```{r results='hold'}
# Microsoft vs T-bills
wM_seq2 <- seq(0, 2, by = 0.1)
mt_df <- tibble(
    w_a = wM_seq2,
    w_b = 1 - w_a,
    mu_p = w_a * mu.vec["Microsoft"] + (1 - w_a) * rf,
    var_p = w_a^2 * var[2],
    sd_p = sqrt(var_p),
    portfolio = "Microsoft-Tbill"
)
cat("Microsoft-Tbill portfolios:\n\n")
mt_df %>%
    as.data.table() %>%
    print(topn = 4)
```


### Plot the mean-standard deviation frontier {-}

```{r}
# combine and prepare for plotting
all_df <- bind_rows(bm_df, bt_df, mt_df)

# Plot frontier
ggplot(all_df, aes(x = sd_p, y = mu_p, color = portfolio)) +
    geom_path() +
    geom_point(size = 1) +
    labs(
        x = TeX("SD($r_p$)"), y = TeX("E($r_p$)"),
        title = "Mean-Standard Deviation Frontier"
    ) +
    scale_color_manual(values = c("Boeing-Microsoft" = "#EE0000FF", "Boeing-Tbill" = "#3B4992FF", "Microsoft-Tbill" = "#008B45FF")) +
    theme_minimal() +
    theme(legend.position = "bottom")
```

### Sharpe Ratios of Individual Assets {-}
```{r}
# Sharpe ratios of individual assets
sharpe_assets <- (mu.vec - rf) / sigma
```

```{r echo=FALSE}
sharpe_assets %>%
    enframe(name = "Asset", value = "Sharpe") %>%
    knitr::kable(digits = 4) %>%
    column_spec(c(1, 2), width = "3cm") 
```

## Global Minimum Variance Portfolio (GMV)

```{r}
# GMV portfolio using IntroCompFinR
gmv.port <- globalMin.portfolio(er = mu.vec, cov.mat = sigma2.mat)
gmv.port
```

```{r results='hold'}
cat("Sharpe ratio of GMV portfolio:\n")
((gmv.port$er - rf) / gmv.port$sd) %>% round(4)
```

## Tangency Portfolio

```{r}
# GMV portfolio using IntroCompFinR
tan.port <- tangency.portfolio(er = mu.vec, cov.mat = sigma2.mat, risk.free = rf)
tan.port
```

```{r results='hold'}
cat("Sharpe ratio of Tangency portfolio:\n")
((tan.port$er - rf) / tan.port$sd) %>% round(4)
```