---
title: "Practical Exercises: Portfolio Performance Evaluation"
output: 
  pdf_document:
    number_sections: true
header-includes:
  - \usepackage[bookmarksnumbered=true]{hyperref}
---

```{r chunk-opt, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6.5, fig.height = 4.5, fig.align = "center")
```

# Packages

Run this once. If a package is already installed, `install.packages()` will
simply reinstall it, so you can comment out the line afterwards.

```{r install, eval=FALSE}
install.packages(c(
  "PerformanceAnalytics",  # every performance measure used below
  "xts",                   # the time series container it expects
  "tidyverse",             # dplyr / ggplot2 / readr
  "knitr", "kableExtra",   # tables
  "scales"                 # percentage formatting
))
```

```{r libs}
library(PerformanceAnalytics)
library(xts)
library(tidyverse)
library(knitr)
library(kableExtra)
library(scales)
```

`PerformanceAnalytics` masks a few base functions (`legend`, `plot`) and
attaches `xts` and `zoo` for you. The masking warnings are harmless.

The figures below use each package's default styling. You do not need any
particular theme -- feel free to restyle them however you like.


# Class Example: An Active ETF against its Benchmark

We evaluate **ARK Innovation (ARKK)**, an actively managed ETF, against
**SPY** (an S&P 500 index fund) over **120 months, 2015--2024**.

Performance evaluation needs exactly three monthly series:

1. the return on the portfolio being evaluated,
2. the return on a benchmark, and
3. the risk-free rate.

## The data {-}

The returns are supplied in `data/etf_ARKK_case_study_monthly.csv`, in the `data/` folder beside this file. Keep the two together, or adjust the path below. It holds 120 monthly rows with four columns:

| Column | Meaning |
| --------- | ----------------------------------------------------------- |
| `Date`    | last calendar day of the month |
| `ptf_ret` | simple return on ARKK, from adjusted prices (dividends and splits included) |
| `bmk_ret` | simple return on SPY, same basis |
| `rf`      | one-month T-bill return, from the Fama-French data library, as a decimal |

```{r read-data}
ret <- read.csv("data/etf_ARKK_case_study_monthly.csv")
ret$Date <- as.Date(ret$Date)
```

Almost every `PerformanceAnalytics` function wants an `xts` object, so convert
once and reuse. Give the columns readable names -- they become the labels in
every table and chart below.

```{r to-xts}
ret_xts <- xts(ret[, c("ptf_ret", "bmk_ret")], order.by = ret$Date)
colnames(ret_xts) <- c("ARK Innovation", "S&P 500")

rf_xts <- xts(ret$rf, order.by = ret$Date)
colnames(rf_xts) <- "rf"   # a name is required, not optional
```

Data preview:

```{r show-data, echo=FALSE}
rbind(head(ret, 5), tail(ret, 5)) %>%
  mutate(across(2:4, ~ percent(.x, accuracy = 0.01))) %>%
  kable(row.names = FALSE, booktabs = TRUE,
        col.names = c("Date", "Portfolio", "Benchmark", "Risk free"),
        ) %>%
  kable_styling(latex_options = "HOLD_position")
```

Note the risk-free rate is *exactly zero* in 2015 and back above 4% a year by
2024. Over a ten-year sample it is not a constant, and treating it as one will
distort every excess-return measure.

## Cumulative performance

`charts.PerformanceSummary` stacks three views: the cumulative return, the
monthly returns, and the drawdown from the running peak.

```{r chart-summary, fig.height=6}
charts.PerformanceSummary(ret_xts, main = "ARKK vs SPY", legend.loc = "topleft")
```

At the 2021 peak the fund was worth **3.7 times** the benchmark; by the end of
2024 it is **behind** it. The drawdown panel shows a fall of about **77%** from
which it has still not recovered.

The drawdown panel is the one investors actually feel, and neither the Sharpe
ratio nor alpha will tell you about it.

## Summary statistics

```{r stats}
stats <- table.Stats(ret_xts, digits = 4)
```

```{r stats-print, echo=FALSE}
stats %>%
  kable(digits = 4, booktabs = TRUE, format = "latex", linesep = "") %>%
  kable_styling(latex_options = "HOLD_position")
```

Things to notice:

- `Kurtosis` is **excess** kurtosis, so positive values mean fat tails.
- The skew runs in opposite directions for the two series.

## Risk-adjusted performance

```{r measures}
# Sharpe ratio. geometric = FALSE reproduces the sqrt(12) scaling from the
# lecture; the default TRUE annualizes the mean geometrically instead.
sharpe <- SharpeRatio.annualized(ret_xts, Rf = rf_xts, geometric = FALSE)

# Everything CAPM-related in one call. scale = 12 because the data is monthly.
capm <- table.CAPM(Ra = ret_xts[, 1], Rb = ret_xts[, 2],
                   Rf = rf_xts$rf, scale = 12, digits = 4)
```

```{r measures-print, echo=FALSE}
rbind(sharpe,
      `Beta`             = c(capm["Beta", ], 1),
      `Annualized Alpha` = c(capm["Annualized Alpha", ], 0),
      `R-squared`        = c(capm["R-squared", ], 1)) %>%
  kable(digits = 3, booktabs = TRUE) %>%
  kable_styling(latex_options = "HOLD_position")

capm[c("Tracking Error", "Active Premium",
       "Information Ratio", "Treynor Ratio"), , drop = FALSE] %>%
  kable(digits = 3, booktabs = TRUE) %>%
  kable_styling(latex_options = "HOLD_position")
```



## Performance by year

`table.AnnualizedReturns` summarises a whole sample. To get one column per
calendar year, split the series and apply it to each piece.

```{r by-year}
table_by_year <- function(R, Rf) {
  yr <- mapply(function(r, rf) table.AnnualizedReturns(r, Rf = rf),
               split(R,  f = "years"),   # xts::split.xts
               split(Rf, f = "years"),
               SIMPLIFY = FALSE)
  out <- do.call(cbind, yr)
  colnames(out) <- unique(format(index(R), "%Y"))
  # cbind keeps the first year's Rf in the Sharpe label, which is misleading
  # once the other columns use their own year's rate
  rownames(out)[3] <- "Annualized Sharpe"
  t(out)
}

performance_by_year <- table_by_year(ret_xts$`ARK Innovation`, rf_xts$rf)
```

```{r by-year-print, echo=FALSE}
performance_by_year %>%
  as.data.frame() %>%
  mutate(across(c(`Annualized Return`, `Annualized Std Dev`),
                ~ percent(.x, accuracy = 0.01)),
         `Annualized Sharpe` = number(`Annualized Sharpe`, accuracy = 0.01)) %>%
  kable(align = "r", booktabs = TRUE, format = "latex", linesep = "") %>%
  kable_styling(latex_options = "HOLD_position")
```


## Rolling performance

A 12-month rolling window shows how the measures move through time instead of
collapsing the decade into one number.

```{r chart-rolling, fig.height=6}
charts.RollingPerformance(R = ret_xts, width = 12, Rf = rf_xts$rf,
                          main = "Rolling 12-month performance",
                          legend.loc = "topleft")
```

## Risk decomposition

```{r risk}
risk <- table.SpecificRisk(Ra = ret_xts[, 1], Rb = ret_xts[, 2],
                           Rf = rf_xts$rf, digits = 4)
```

```{r risk-print, echo=FALSE}
risk %>%
  kable(digits = 3, booktabs = TRUE) %>%
  kable_styling(latex_options = "HOLD_position")
```

\newpage

# Practical Exercise: A Second Fund

A second dataset, `data/utilities_fund_monthly.csv`, has the same four columns
and covers the same 120 months. The benchmark and risk-free columns are
identical to the class example; only `ptf_ret` differs -- it is a **utilities
sector fund**, a defensive, low-beta portfolio, and about as far from ARKK as
you can get while still holding US equities.

```{r exercise-data}
ex <- read.csv("data/utilities_fund_monthly.csv")
ex$Date <- as.Date(ex$Date)
rbind(ex %>% head(5), ex %>% tail(5)) %>% 
  mutate(across(2:4, ~ percent(.x, accuracy = 0.01))) %>%
  kable(digits = 4, row.names = FALSE, format = "latex", booktabs = TRUE,
        col.names = c("Date", "Portfolio", "Benchmark", "Risk free")) %>% 
  kable_styling(latex_options = "HOLD_position")

ex_xts <- xts(ex[, c("ptf_ret", "bmk_ret")], order.by = ex$Date)
colnames(ex_xts) <- c("Utilities Fund", "S&P 500")

ex_rf <- xts(ex$rf, order.by = ex$Date)
colnames(ex_rf) <- "rf"
```

Work through the same steps as the class example and answer the following.

(a) Plot the cumulative performance. When was the fund ahead of its benchmark,
    and what was its worst drawdown? How does that drawdown compare with
    ARKK's $77\%$?

(b) Report the summary statistics for the fund and the benchmark monthly returns using `table.Stats()`. How do the two series compare in terms of mean, volatility, skew and kurtosis?

(c) Compute the annualized Sharpe ratio for the fund and the benchmark. Which
    would you rather hold as your *entire* risky portfolio?

(d) Compute the beta, Jensen's alpha, the Treynor ratio and the information
    ratio.

(e) Decompose total risk into systematic and specific risk. What fraction of
    the fund's variance is firm-specific? Compare the $R^2$ with ARKK's
    $0.52$: which fund is the market explaining more of?

(f) On the evidence, would you call this fund's manager skilled? 
